### 2017-01-29 manuscript.rmd
changes to text

The reason that paralog pairs have Tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are different rates of Tau evolution following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. Closely related paralogs, for example, are expected to be more similar than distantly related orthologs [@Studer:2009bm]. This has several implications. It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and annotations of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj]. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study. For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches.

@@

Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. This has important biological implications. It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought. It also has technical implications. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs. 
There has been considerable recent intrest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi]. Our reanalysys is consistent with multiple studies that have not found support for the ortholog conjecture in comparative analyses of gene ontology [@Studer:2009bm; @Nehrt:2011bm] or gene expression [@Yanai:2004hk]. A previous study of expression concluded that there was strong support for the ortholog conjecture , though like KMRR it used pairwise comparisons.

; @Altenhoff:2012ea

>>>>

The reason that paralog pairs have Tau values that are more different than ortholog pairs is that they tend to be more distantly related, not because there are different rates of Tau evolution following duplication and speciation events. This means that having information on whether two genes are orthologs or paralogs provides little added information about expression beyond knowing how distantly related the two genes are. This has several implications. It is an example of the limitations of the concepts of orthology and paralogy [@Dunn:2016er]. These terms can have straightforward meaning in small gene trees with simple duplication/speciation histories, but the utility of the terms breaks down on larger more complex gene trees. Orthology and paralogy are annotations on the tips of the tree that are derived from the inferred structure of the tree and history of duplication and speciation at internal tree nodes. In this sense, orthology and paralogy are statements about the internals of the tree that are distilled into statements at the tips of the tree. Much is lost in the process, though. These are the same pre-computed trees we use in our phylogenetic re-analyses of their study. For most questions it is much more direct to focus on the structure of the tree and the inferred processes within the tree, such as which internal nodes are duplication or speciation events and how much change occurs along the branches. In this case, KMRR [@KryuchkovaMostacci:2016iw] relied on orthology and paralogy annotations that are derived from the inferred duplication and speciation history on gene phylogenies [@Herrero:2016jj].

@@

There has been considerable recent intrest in, and controversy about, the ortholog conjecture [@Koonin:2005fp; @Studer:2009bm; @Nehrt:2011bm; @Gabaldon:2013fi]. Our results are consistent with multiple studies that have not found support for the ortholog conjecture [@Studer:2009bm; @Nehrt:2011bm; @Yanai:2004hk]. Another sudy found only slightly greater similarity between orthologs than paralogs [@Altenhoff:2012ea].

Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. This has important biological implications. It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as sometimes proposed. It also has technical implications. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs [@Studer:2009bm].

The ortholog conjecture does not have to be an all or nothing question. It may be the case that the rates of phenotypic evolution following duplication may be greater than that following duplication in some organisms, gene families, and evolutionary processes. We just don't find evidence for it when summarizing gene expression acriss tissues with Tau in these mammals. This calls into question the general predictive power of the ortholog conjecture, and until these processes are better understood it will be necessary to test for it in each situation. These tests should be articulated in terms of clear alternative hypotheses [@Studer:2009bm] that make distinct phylogenetic comparative predictions.
